The role of standing variation in parallel adaptation

### 2014-09-17 standing-variation-paper.tex

Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutatoin at different locations within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Three G6PD deficiency alleles that also confer resistance to malaria 
are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
\plr{what are Ruwende's results? feels a bit vague here}
Given such a strong pressure on these alleles they should have risen quickly to fixation, 
so their presence at intermediate frequency over a broad geographic area
makes them good candidates for recently balanced polymorphisms. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, 
notably those present in fava beans.

Note that although originally the ``wave of advance'' was presented as the spread of an unconditionally favorable allele
\citep{fisher1936,KPP},
the same model holds in the case of heterozygote advantage \citep{aronson1975nonlinear}, 
with the selected allele spreading locally to the equilibrium frequency (rather than fixation),
with speed determined by the advantage of heterozygotes when rare.

>>>>

Over the past ten thousand years, malaria resistance alleles have arisen at a number of genes 
and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to be examples of convergent adaptation, 
as different derived mutations in the same gene are seen in different individuals.
For example, 
a number of changes have been observed in the $\beta$-globin gene that confer malaria resistance;
and the sickle cell allele may plausibly have arisen by up to five independent occurrences
of the same base pair mutatoin at different locations within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the X-linked G6PD gene, 
with upward of 50 polymorphic variants ($>1\%$ local frequency) having so far been described 
that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. 
These alleles are now found at a combined frequency of around 8\% frequency in malaria endemic areas,
rarely exceeding 20\% \citep{Howes-g6pd-preval}. 
Three G6PD deficiency alleles that also confer resistance to malaria 
are particularly common and relatively well studied: 
the $A-$ allele found in much of sub-Saharan Africa; 
the Med allele found in the Mediterranean and Middle East; 
and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd, Louicharoen-g6pd}. 
Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread over the past 
few thousand years  \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other known malaria resistance alleles. 
Population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the $A-$ allele. 
\plr{what are Ruwende's results? feels a bit vague here} \\

Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans (although, It is unclear if this alone is enough of a
selection pressure to maintain the allele in the face of its
considerable advantage). 
The wave of advance works in the case of heterozygote advantage
\citep{aronson1975nonlinear}, with the selected allele spreading
locally to the equilibrium frequency (rather than fixation). As such
our framework is applicable to the spread of G6PD, with speed determined by the advantage of heterozygotes when rare.
We assume that before malaria became prevalent, the G6PD deficiency alleles suffered a
decrease in relative fitness of $s_d$ in heterozygote and homozygotes females and
hemizygote male. Assuming that the underlying causes of this drop in fitness have
not changed, we estimate that $s_d$ has to be roughly
XXX to XXX, in order to have resulted in the equilibrium frequency
seen today.
